Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-11-17 draft_lm1.tex

The estimation of the  number of susceptibles have been attempted 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

\subsection*{Estimating the reproductive number}

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
Also, although there probably are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

For the fitting procedure the time series was divided into $J=10$ epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.
Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

@@

% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.

>>>>

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

In this section we will start by describing how we estimated the effective 
reproductive number, $R_t$, relative to the global dengue cases, and it's 
posterior probability distribution. 
We then proceed to describe the SIR model 
used to represent the global incidence and how the $R_t$ is integrated into the 
model to allow for time varying transmission. 
Then we describe the Bayesian 
estimation methodology applied to find the posterior probability distributions 
of the susceptibles to the main circulating Dengue viruses for each epidemic, 
using Dengue incidence time series for Rio de Janeiro from 1996 to 2014.
In the last subsection we describe how we estimated the Attack ratio of each 
epidemic from the estimated susceptibility and observed incidence.

\subsection*{Estimating the effective reproductive number, $R_t$}

@@

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number as defined 
in~\ref{eq:Rtestimate}.
We parameterize our model with the varying transmission using the system of 
ordinary differential equations
\begin{align}
 \frac{dS}{dt} &= -\beta(t)SI \\
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\
 \frac{dR}{dt} &= \tau I&
\end{align}
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
This is a rather simplified model, in which, for instance, the vector is left 
out.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
does not discriminate between them due to the lack of serotype-specific data.

@@

For the fitting procedure the incidence time series was divided into $J=13$ 
epidemic 
windows that corresponded to significant raises in incidence and normalized to 
lie on the $[0,1]$ interval.
Also, as the numerical integration routine needs to treat $\beta(t)$ as a
continuous function of time because of the variable step size, we used linear 
interpolation to obtain values of $R_t$ for any time point.

@@

\subsection*{Calculating the attack ratio}

The attack ratio of an epidemic is define by the number of infections divided 
by the size of the population at risk.

\begin{equation}
\label{eq:AR}
A=\frac{cases}{Pop.\, at\, risk} 
\end{equation}

Based on the what have been discussed so far, we can rewrite (\ref{eq:AR}) for 
each epidemic $j$ as
\begin{equation}
\label{eq:AR2}
 A_{j}=\frac{\sum Y_j}{S_{0j}}
\end{equation}
where $S_{0j}$ is the number of susceptibles before each epidemic $j$, 
estimated as described above.

% Results and Discussion can be combined.
\section*{Results and Discussion}

The time series used for all the analyses consist of officially notified dengue 
cases agregated weekly.
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$. The authors are also grateful to Claudia T. 
Codeco for helpful discussions about the manuscript.
